Evidence for a Color Parameter Unassociated With Dust Within the Type~Ia Supernovae of the Nearby Supernova Factory

### 2017-09-07 apj.tex
more changes

\begin{abstract}
Through empirical modeling of its observed signal, the peak absolute magnitude of a Type~Ia supernova (SN~Ia) can be accurately determined,
making SNe~Ia  excellent distance indicators.  Improved modeling of SN~Ia absolute magnitudes can account for
further physical
diversity that is expected but not included
in current models, and so consequently can lead to more precise per-object distances with smaller systematic uncertainties.  In this article, we present   
an empirical model for SN~Ia peak magnitudes with two color parameters and dependence on the equivalent widths of CaII, SiII, and SiII velocity
as applied to the supernova sample of the Nearby Supernova Factory.  The peak magnitudes in synthetic
broadband photometry and their colors are found to be 
dependent on the spectral equivalent widths and the color parameters, to better than 
%-----
$(1-5\times10^{-5})$
%------
confidence.
The color changes allowed by the model are similar to those allowed by the extinction model of  \citet{1999PASP..111...63F} with a free
total-to-selective extinction parameter $R^F$.  As an ensemble, the supernova sample has an effective
%-----
$R^F = 2.44$.
%-----
Our model
 is able to reproduce the optical colors of the highly reddened SN~2014J better than the \citet{1999PASP..111...63F} dust model.
Inspired that with no conditioning our  two-parameter model produces behavior similar to that expected from Galactic dust,
we extend  the model to three color parameters.
In this new case we obtain consistent results for
all the parameters in common,
plus a detection of a third parameter at
%--- table.py
$\sim 90\%$
%-----
confidence;
this new parameter represents supernova color diversity
beyond that tracked by the spectral features and two-parameter dust models.
\end{abstract}

\keywords{supernovae: general; supernovae: SN 2014J}

@@

In this article we use the idea that spectral indicators carry information on intrinsic supernova colors at peak magnitude.
This approach is taken by \citet{2011A&A...529L...4C}, who find that after standardization based on Ca and Si features, remaining residual color
variation is consistent with Milky Way dust models.
We accommodate up to  three independent color parameters: the model is constructed such that 
two are assigned as extrinsic processes attributed to dust, and one is attributed to intrinsic supernova
properties.  (We subsequently find that extrinsic and intrinsic properties cannot be distinguished with our data.)
The data used in this analysis are described in \S\ref{data:sec}.
In \S\ref{model:sec} we present a
first analysis with two extrinsic parameters and omitting the intrinsic parameter.  The properties of the two extrinsic parameters
are consistent with the expectations of parameterized Milky Way extinction models, and do an excellent job of fitting the data of the out-of-sample
supernova SN2014J that has an extremely low total-to-selective extinction.
In \S\ref{model2:sec} we add the intrinsic parameter in the analysis, and find strong evidence for its influence on observed magnitudes.
We summarize our findings, and relate our results to light-curve shape, SiII velocity, and the host-galaxy global stellar ``mass step'' in SN~Ia Hubble
residuals in \S\ref{discussion:sec}.

@@

The Model~II fit of SN~2014J \citet{2014ApJ...788L..21A}  data 
gives $k_0$ and $k_1$ that are qualitatively similar to  those from the fit to Model~I but with much larger uncertainties.
These parameters are now highly correlated with $D$, which itself is consistent with zero.  The results resemble those of
Model~I shown in Figure~\ref{sn2014j:fig}, except for larger error bars for the color excesses projected onto the dust model.

@@

Evidence for the intrinsic origin of $\pmb{\delta}D$ is seen in the relation between observed magnitudes corrected for intrinsic magnitude correlations with
spectral features (but not for the color parameters)
and observed  color.  Figure~\ref{speccorr:fig} shows such a plot for corrected observed $B_o$ magnitudes
as a function observed $B_o-V_o$, in correspondence with Figure~3 of  \citet{2017ApJ...842...93M}.  Supernovae with extreme blue colors exhibit a shallower slope than those
with extreme red colors.   \citet{2017ApJ...842...93M} have demonstrated that this is the expected behavior for
dust with constant $\beta$ (the SALT2 equivalent of $R$) with an exponential extinction distribution and an intrinsic source of color variablilty
with a smaller value of $\beta$.  Without interstellar dust, supernovae are bluer and take the intrinsic $\beta$ slope for the color-magnitude
relation.  When the intrinsic contribution to color is dwarfed by dust, supernovae are bluer and take the larger dust $\beta$ slope.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5.2in]{output25/speccorr_color.pdf}
      \caption{
      \color{red}Observed $B$-band magnitude corrected for intrinsic magnitude correlations with spectral features as a function of observed color $B_o-V_o$.
      The two solid lines are the linear regressions for points with abscissa values less than and greater than zero.
	\color{black}
   \label{speccorr:fig}}
\end{figure}

In contrast, if the colors were due to a single physical dust described by three parameters,
the observed correlation between the color-magnitude slope and observed color seen in Figure~\ref{speccorr:fig}
would manifest itself in correlation between the $D$,
$k_0$ and $k_1$ parameters.  To the contrary,
the correlations between these parameters  are consistent with zero as given  in Eqn.~\ref{corr2p:eqn}.

If  $\delta D$  is intrinsic, it can be attributed to an intrinsic supernova property  or a dust not described by $\gamma k$.
To first order $\pmb{\delta}$ behaves like standard dust;   $>98$\% of its length projects onto the expectation of  \citetalias{1999PASP..111...63F}
with constant $R^F$.
Therefore, a straightforward explanation is that it describes the absorption due to some form of dust
that does not exactly obey the  \citetalias{1999PASP..111...63F} model.
The departure from   \citetalias{1999PASP..111...63F}
can be seen qualitatively; the decrease in magnitude change toward redder wavelengths, seen in Figure~\ref{deltaratio:fig}, matches the 
expectation for the circumstellar form of dust.
However, we
lack the signal-to-noise to show consistency with this explanation.
The best-fit elements $\delta_X$ do not all share the same sign, as would have been expected for dust.
Confining ourselves to the 98\% of   $\pmb{\delta}$ that lies on the dust plane, associating 
$D$ with $A_V$ would imply an unphysical $R^F=0.5$.
We do not have conclusive evidence that  $\delta$ is being a property of a different form of dust, rather
than representing intrinsic supernova color variation.
\color{black}

\subsection{Conclusion}
To summarize, we model SNe~Ia broadband optical peak magnitudes allowing for correlations with spectral features at peak and
extrinsic color parameters.  Analyzing SNfactory data with this model, we find significant evidence that the above parameters do
affect supernova magnitudes and colors.  This model  does an excellent job in
describing SN~2014J, a supernova with extreme colors that was not used in the training.  
This work represents the first determination of the dust extinction curve outside the Local Group
derived entirely independently of assumptions about the shape of the extinction curve and/or assumptions about the
distribution of $A_V$.  
Nesting this model in an extended model produces a result that 
gives the first identification of  a new parameter that affects supernova
colors in a manner that is distinct from the expectations from dust or the spectral features $EW_{Ca}$, $EW_{Si}$, and $\lambda_{Si}$.

>>>>

\begin{abstract}
Through empirical modeling of its observed signal, the peak absolute magnitude of a Type~Ia supernova (SN~Ia) can be accurately determined,
making SNe~Ia  excellent distance indicators.  Improved modeling of SN~Ia absolute magnitudes can account for
further physical
diversity that is expected but not included
in current models, and so consequently can lead to more precise per-object distances with smaller systematic uncertainties.  In this article, we present   
an empirical model for SN~Ia peak magnitudes with two color parameters and dependence on the equivalent widths of CaII, SiII, and SiII velocity
as applied to the supernova sample of the Nearby Supernova Factory.  The peak magnitudes in synthetic
broadband photometry and their colors are found to be 
dependent on the spectral equivalent widths and the color parameters, to better than 
%-----
$(1-5\times10^{-5})$
%------
confidence.
The color changes allowed by the model are similar to those allowed by the extinction model of  \citet{1999PASP..111...63F} with a free
total-to-selective extinction parameter $R^F$.  As an ensemble, the supernova sample has an effective
%-----
$R^F = 2.44$.
%-----
%Our model
% is able to reproduce the optical colors of the highly reddened SN~2014J better than the \citet{1999PASP..111...63F} dust model.
Inspired that with no conditioning our  two-parameter model produces behavior similar to that expected from Galactic dust,
we extend  the model to three color parameters.
In this new case we obtain consistent results for
all the parameters in common,
plus a detection of a third parameter at
%--- table.py
$\sim 90\%$
%-----
confidence;
this new parameter represents supernova color diversity
beyond that tracked by the spectral features and two-parameter dust models.
\end{abstract}

\keywords{supernovae: general; supernovae}

@@

In this article we use the idea that spectral indicators carry information on intrinsic supernova colors at peak magnitude.
This approach is taken by \citet{2011A&A...529L...4C}, who find that after standardization based on Ca and Si features, remaining residual color
variation is consistent with Milky Way dust models.
We accommodate up to  three independent color parameters: the model is constructed such that 
two are assigned as extrinsic processes attributed to dust, and one is attributed to intrinsic supernova
properties.  (We subsequently find that extrinsic and intrinsic properties cannot be distinguished with our data.)
The data used in this analysis are described in \S\ref{data:sec}.
In \S\ref{model:sec} we present a
first analysis with two extrinsic parameters and omitting the intrinsic parameter.  The properties of the two extrinsic parameters
are consistent with the expectations of parameterized Milky Way extinction models.
%, and do an excellent job of fitting the data of the out-of-sample
%supernova SN2014J that has an extremely low total-to-selective extinction.
In \S\ref{model2:sec} we add the intrinsic parameter in the analysis, and find strong evidence for its influence on observed magnitudes.
We summarize our findings, and relate our results to light-curve shape, SiII velocity, and the host-galaxy global stellar ``mass step'' in SN~Ia Hubble
residuals in \S\ref{discussion:sec}.

@@

%The Model~II fit of SN~2014J \citet{2014ApJ...788L..21A}  data 
%gives $k_0$ and $k_1$ that are qualitatively similar to  those from the fit to Model~I but with much larger uncertainties.
%These parameters are now highly correlated with $D$, which itself is consistent with zero.  The results resemble those of
%Model~I shown in Figure~\ref{sn2014j:fig}, except for larger error bars for the color excesses projected onto the dust model.

@@

Color-based magnitude corrections as a function of color excess provide a clue to ``intrinsic'' and ``dust'' attributions of
the combined effect of the color terms $k_0$, $k_1$, and $D$.  
All supernovae experience  intrinsic color variation.
Only a subset encounter dust:
geometric projection effects lead to the expectation for a subset of supernovae having zero
dust column density and the
complement subset having an approximate exponential distribution \citep{1998ApJ...502..177H}.
Differences between the expected distributions of
distinct intrinsic effects and dust lead to the prediction of a non-constant slope for the relationship between
total (from both intrinsic and dust)  magnitude corrections and color excess.

\citet{2017ApJ...842...93M} present an instructive simulation, producing magnitudes and colors of a set of
supernovae 
with an intrinsic source of color variability having relatively smaller value of $\beta$
and a realistic distribution of dust extinction having a relatively larger $\beta$ value.  The expectation is for
supernovae encountering no interstellar dust  to tend to be bluer and exhibit the intrinsic $\beta$ slope for the color-magnitude
relation.  Supernovae encountering thick dust tend to be redder and exhibit the larger dust $\beta$ slope.
The expectation is borne out in their Figure 3, which shows 
observed color versus light-curve shape corrected magnitudes, and is seen in their data Figure 6.
The distinctive expectations manifest in the plot  for the above model allow
the authors to use  information inherent in the distributions to make distinct measurements of intrinsic and dust slopes (referred
therein as $\beta$ using the terminology of SALT2),
despite being
restricted to using one color observable that is  affected by two sources of extinction.

Our Model~II has for each supernova three color parameters that can contain information on more
than one source of color variablility. This allows us to construct from our fit results per-supernova magnitude corrections as a function
of color, to produce a model-based figure analogous to Figure~3 of \citet{2017ApJ...842...93M}.
Such a result is shown in Figure~\ref{speccorr:fig}, which plots
$U$-band magnitude corrections from the three color parameters $\gamma^0_{U} k_0 +\gamma^1_{U} k_1 + \delta_U D$
as a function of $E_o(B-V)$, where the expected color accounts for spectral-feature corrections.
The shallower slope at blue color transitioning to a larger slope at red color 
is correspondence with the expectation of a mixed intrinsic--dust model of the form considered by  \citet{2017ApJ...842...93M}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output25/ccorr1e.pdf}
   \includegraphics[width=2.8in]{output25/ccorr2e.pdf}
      \caption{
      \color{red}
      The $U$-band magnitude corrections from the three color parameters $\gamma^0_{U} k_0 +\gamma^1_{U} k_1 + \delta_U D$
are plotted as a function of $E_o(B-V)$, where the expected color accounts for spectral-feature corrections.  The left plot shows the full
range of values whereas the right zooms in the upper-left corner of the population.
	\color{black}
   \label{speccorr:fig}}
\end{figure}

If  $\delta D$  is intrinsic, it can be attributed to an intrinsic supernova property  or a dust not described by $\gamma k$.
To first order $\pmb{\delta}$ behaves like standard dust;   $>98$\% of its length projects onto the expectation of  \citetalias{1999PASP..111...63F}
with constant $R^F$.
Therefore, a straightforward explanation is that it describes the absorption due to some form of dust
that does not exactly obey the  \citetalias{1999PASP..111...63F} model.
The departure from   \citetalias{1999PASP..111...63F}
can be seen qualitatively; the decrease in magnitude change toward redder wavelengths, seen in Figure~\ref{deltaratio:fig}, matches the 
expectation for the circumstellar form of dust.
However, we
lack the signal-to-noise to show consistency with this explanation.
The best-fit elements $\delta_X$ do not all share the same sign, as would have been expected for dust.
Confining ourselves to the 98\% of   $\pmb{\delta}$ that lies on the dust plane, associating 
$D$ with $A_V$ would imply an unphysical $R^F=0.5$.
We do not have conclusive evidence that  $\delta$ is being a property of dust distinct from that described by $\gamma$, rather
than representing intrinsic supernova color variation.
\color{black}

\subsection{Conclusion}
To summarize, we model SNe~Ia broadband optical peak magnitudes allowing for correlations with spectral features at peak and
extrinsic color parameters.  Analyzing SNfactory data with this model, we find significant evidence that the above parameters do
affect supernova magnitudes and colors.  
%This model  does an excellent job in
%describing SN~2014J, a supernova with extreme colors that was not used in the training.  
This work represents the first determination of the dust extinction curve outside the Local Group
derived entirely independently of assumptions about the shape of the extinction curve and/or assumptions about the
distribution of $A_V$.  
Nesting this model in an extended model produces a result that 
gives the first identification of  a new parameter that affects supernova
colors in a manner that is distinct from the expectations from dust or the spectral features $EW_{Ca}$, $EW_{Si}$, and $\lambda_{Si}$.
